---
title: "Using fhicolours"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using fhicolours}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 3.5,
  fig.align = "center",
  dpi = 96
)
```

The **fhicolours** package provides scales for **ggplot2** aligned with Folkehelseinstituttet (FHI) palette guidance for figures and statistics.

According to [**FHIs fargepalett**](https://www.fhi.no/om/fhi/visuell-profil/innhold/farger/), the overall visual palette includes **hovedfarger** (main blues, reds, greens, yellows), **gråfarger** (including gråblå and greige tones that work alongside the main colours), **ekstrafarger**, backgrounds, and a dedicated set of colours for **statistical visualization** (graphs, maps, diagrams). FHI applies these across fhi.no, social media, and other channels.

The statistics-oriented advice on that page boils down to three patterns that this package mirrors:

- **Quantitative (sequential)** ramps are built to keep contrast comparable across hues.
- **Qualitative** colours stress lightness contrast to help meet accessibility expectations (WCAG and APCA). They are meant for categories **without** intrinsic order—examples FHI mentions include country, ethnicity, sex, or political party.
- **Diverging** palettes suit numeric data with a **neutral centre**; FHI notes they can be extended using additional tones from **B2** and **R1**.

FHI also specifies institutional **text** colour **`#222222`** for a softer alternative to pure black.

```{r setup}
library(ggplot2)
library(fhicolours)
```

## Qualitative palettes

Use `scale_color_fhi()` or `scale_fill_fhi()` with the default `discrete = TRUE` when the aesthetic maps to a factor or character variable. The `qualitative` palette follows the sixteen-colour ordered list under *Kvalitativ fargepalett* on the FHI page (ggplot interpolates if you need more).

```{r qualitative}
ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point(alpha = 0.85, size = 2) +
  scale_color_fhi(palette = "qualitative") +
  theme_minimal() +
  labs(title = "Discrete colour: qualitative palette")
```

## Quantitative palettes

Sequential palettes such as `b2`, `r1`, `gr1`, and `gu1` are ordered ramps from FHI’s quantitative palettes. **`grable`** and **`greige`** are the gråblå and greige scales from the profile’s grey section—intended to pair with the main brand hues. For a continuous variable set `discrete = FALSE` so ggplot uses a smooth gradient (`scale_*_gradientn()` under the hood).

```{r quantitative}
grid <- expand.grid(x = seq_len(24), y = seq_len(24))
grid$value <- with(grid, sin(x / 4) + cos(y / 4))

ggplot(grid, aes(x, y, fill = value)) +
  geom_raster() +
  scale_fill_fhi(palette = "b2", discrete = FALSE) +
  coord_equal() +
  theme_minimal() +
  labs(title = "Continuous fill: sequential B2")
```

## Diverging palettes

`diverging_b2_r1` combines **R1** and **B2**-family tones with white at the centre, in line with FHI’s diverging guidance (two hues with a neutral midpoint, extendable with more B2/R1 steps). Use it when the middle of the scale should read as neutral—for example deviations from zero or a reference value.

```{r diverging}
grid2 <- expand.grid(
  x = seq(-pi, pi, length.out = 40),
  y = seq(-pi, pi, length.out = 40)
)
grid2$value <- with(grid2, sin(x) * cos(y))

ggplot(grid2, aes(x, y, fill = value)) +
  geom_raster() +
  scale_fill_fhi(palette = "diverging_b2_r1", discrete = FALSE) +
  coord_equal() +
  theme_minimal() +
  labs(title = "Continuous fill: diverging B2 / R1")
```

## Matching scales

Use `scale_fill_fhi()` when colour is mapped to `fill` (tiles, bars, densities) and `scale_color_fhi()` when it is mapped to `colour` (points, lines). The same `palette` and `discrete` arguments apply to both.

```{r fill-bars, fig.height = 4}
ggplot(mpg, aes(class, fill = drv)) +
  geom_bar(position = "fill") +
  scale_fill_fhi(palette = "qualitative") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1)) +
  labs(title = "Discrete fill: qualitative palette", y = NULL)
```

To reverse any ramp (for example so darker means “more”), pass `reverse = TRUE` to the scale.
